\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2024-05-28 doc/paper-jss/sdmTMB-paper.Rnw
Revisions

@@

Here, we introduce the \R\ package \sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using \TMB\ for model fitting and \pkg{fmesher} or \INLA\ to set up SPDE matrices.
Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
Many individual features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date this full suite of useful features has not been integrated into a single package.
One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \TMB: \sdmTMB), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and approximated in discrete time-intervals.

@@

A GLMM with spatial Gaussian random fields can be written as
\[
\begin{aligned}
\mathbb{E}[y_{\bm{s}}] &= \mu_{\bm{s}},\\
\mu_{\bm{s}} &=
g^{-1} \left( \eta_{\bm{s}} \right),\\
\eta_{\bm{s}} &= \bm{X}_{\bm{s}} \bm{\beta} + \omega_{\bm{s}},\\
\end{aligned}
\]
where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in space $\bm{s}$ is equal to mean $\mu_{\bm{s}}$.
That mean $\mu_{\bm{s}}$ is the result of an inverse link function $g^{-1}$ applied to a linear predictor $\eta_{\bm{s}}$.
In this case, that linear predictor is the combination of a model matrix $\bm{X}_{\bm{s}}$ multiplied by a vector of coefficients $\bm{\beta}$ and a value from a spatial random field $\omega_{\bm{s}}$.
This spatial random field represents the effect of latent spatial variables that are not otherwise accounted for in the model.
Alternatively, $\omega_{\bm{s}}$ can be thought of representing spatially correlated noise arising from unmodeled processes.
More simply, the vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) represents a ``wiggly'' surface with an expected value of zero that is added to the linear predictor in link space (e.g., Figure~\ref{fig:matern-range}).
The vector $\bm{\omega}$ is assumed drawn from a multivariate normal distribution with a covariance matrix $\boldsymbol{\Sigma}_\omega$,
\[
\boldsymbol{\omega} \sim \operatorname{MVNormal}
  \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),
\]
constrained by some function that defines the rate at which spatial covariance decays with distance.

@@

Various covariance functions are possible, but a popular and flexible one is the Mat\'ern \citep{whittle1954, matern1986} (Figure~\ref{fig:matern-range}).
If we define $\lVert h \rVert$ as the Euclidean distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the Mat\'ern covariance $\Phi$ as
\[
\Phi\left( \bm{s}_j, \bm{s}_k \right) =
  \frac{\sigma_m^2}{2^{\nu - 1}\Gamma(\nu)}
  \left(\kappa \lVert h \rVert \right)^\nu K_\nu
  \left( \kappa \lVert h \rVert \right).
\]
The parameter $\sigma_m^2$ is the marginal variance (magnitude of the random field ``wiggles''), $\Gamma$ represents the Gamma function, $K_\nu$ represents the modified Bessel function of the second kind, and $\kappa$ represents the spatial decorrelation rate.
The parameter $\nu$ controls the smoothness of the covariance function.
In practice, $\nu$ is challenging to estimate and herein is fixed at $\nu = 1$ \citep{lindgren2011}.
A more interpretable derived parameter than the spatial decorrelation rate $\kappa$ is the spatial range---a distance at which two points are effectively independent.
A common definition is $\sqrt{8 \nu} / \kappa$ (so if $\nu = 1$, $\mathrm{range} = \sqrt{8} / \kappa$), which has the empirically derived property of the distance at which correlation decays to $\approx 0.13$ \citep{lindgren2011} (Figure~\ref{fig:matern-range}).

@@

In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or adding random intercepts by grouping factors (Figure~\ref{fig:diagram}).
All these components are additive in link space.
Adopting the notation ``main'' for main effects, ``tvc'' for time-varying coefficients, and ``svc'' for spatially varying coefficients, a more complex model could be written as
\[
\begin{aligned}
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma_t} +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\alpha_g +
O_{\bm{s},t} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right).
\end{aligned}
\]
where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector of coefficients that are constrained to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a random field, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O_{\bm{s},t}$ represents an offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}).
We demonstrate these model components in Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}.
% TODO: add MVN(0, sigma)s

@@

So far we have described models with one linear predictor and one family (common terminology in \R\ packages for the combination of an observation likelihood and link).
Frequently, data are better represented with two-part ``delta'' or ``hurdle'' models, which include linear predictors and observation distributions for two processes: zero vs.\ non-zero values and positive values, respectively \citep{aitchison1955}.
Here, we describe two types of delta models, dropping the space and time subscripts for simplicity.
% and use $\bm{X_1} \bm{\beta_1}$ and $\bm{X_2} \bm{\beta_2}$ to refer to generic model matrices and parameter vectors for the first and second model components.

@@

Once the user makes a call to \fct{sdmTMB}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
If needed, data structures required to implement penalized smoothers are constructed with \fct{smooth2random} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
\fct{sdmTMB} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} as interface to the \pkg{Eigen} library in \proglang{C++}.
% TODO Jim:  I recommend adding a citation to package Matrix ( Bates D, Maechler M, Jagan M (2023). _Matrix: Sparse and Dense Matrix Classes and Methods_. R package version 1.6-4, <https://CRAN.R-project.org/package=Matrix>.)), as central dependency of TMB, given the detailed description of the dependency stack. If really motivated, you could also add the reference for Eigen (Guennebaud G, Jacob B, et al. (2010). “Eigen v3.” http://eigen.tuxfamily.org), although it's less clear what citation to use.
The negative marginal log likelihood ivs minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \Rev{A6} \fct{stats::optimHess} \citep{gay1990, r2021}.
Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
it is also possible to apply an ``epsilon'' estimator \citep{thorson2016bias}, which corrects for bias arising from the variance and skewness caused by non-linear transformations of random effects (see Section \ref{index-bias-correct} for an example).
Standard errors on all parameters and derived quantities---including those involving random effects---are calculated using the generalized delta method \citep{kristensen2016, zheng2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to sample from the joint posterior distribution for Bayesian inference.

@@

Prediction on fitted or new data is accomplished with \fct{predict} (\code{?predict.sdmTMB}).
In this paragraph, we include relevant \fct{predict} arguments in parentheses.
The predict method is flexible and includes the option to specify a new data frame (\code{newdata}),
whether to return predictions on the link or response scale (type),
whether to return standard errors (\code{se\_fit},
which can be slow if conditioned on random fields),
whether to condition on the random fields (\code{re\_form}),
whether to condition on the random intercepts (\code{re\_form\_iid}),
which delta model linear predictor to use (\code{model}),
whether to take draws from the joint parameter precision matrix (\code{nsim}),
and whether to use MCMC samples from a \pkg{tmbstan} model fit (\code{mcmc\_samples}).

A variety of model evaluation tools are available.
A \fct{residuals} method calculates various types of residuals.
The default is a form of randomized quantile residuals \citep{dunn1996}.
New observations can be simulated from a fitted model with \fct{simulate} or new data can be simulated without a fitted model (``de novo'') with \fct{sdmTMB\_simulate}.
\fct{sdmTMB\_cv} facilitates cross validation.

@@

<<remotes, eval=FALSE, echo=TRUE>>=
install.packages("remotes")
remotes::install_github("pbs-assess/sdmTMB", dependencies = TRUE)
@

@@

Users willing to replace the default \R\ \pkg{BLAS} \citep{blas}
% and \pkg{LAPACK} \citep{lapack}
library
with an optimized version \citep[e.g., \pkg{openBLAS};][]{openblas}
can expect about an order of magnitude increase in model fitting speed.
Suggestions are included in the package README.

@@

We can visualize the mesh object with the associated plotting method (Figure~\ref{fig:dog-binomial-mesh2}).
Our mesh has 563 (\code{mesh\_pcod2\$mesh\$n}) vertices.
Usually, mesh complexity has the single largest influence on the speed of fitting these models.

@@

<<morans, echo=TRUE>>=
inv_dist_matrix <- 1 / as.matrix(dist(pcod[,c("X", "Y"), ]))
diag(inv_dist_matrix) <- 0
set.seed(1)
r_bin <- residuals(fit_bin, type = "mle-mvn")
set.seed(1)
r_bin_rf <- residuals(fit_bin_rf, type = "mle-mvn")

ape::Moran.I(r_bin, weight = inv_dist_matrix)$p.value
ape::Moran.I(r_bin_rf, weight = inv_dist_matrix)$p.value
@

% We omit example code for Moran's I calculation for brevity but include it in the code supplement.
We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but a lack of evidence for the model with a random field (p $>$ 0.01) suggesting that residual spatial autocorrelation is alleviated by including the random field.
We note that the specific p-value is dependent on the seed due to the randomization in the randomized quantile residuals \citep{dunn1996}.

@@

\Rev{A1}We note that caution is warranted in performing model selection via marginal AIC for models involving penalized smoothing (spatial or otherwise) \citep[e.g.][]{greven2010, safken2021}.
In marginal AIC calculation, the degrees of freedom is based on the number of fixed effects and does not account for the degree of random effect penalization as conditional AIC would.
Methods to estimate effective degrees of freedom were recently demonstrated in \citep{thorsonXX}, but are not yet included in \sdmTMB.

@@

As an alternative to AIC, we can conduct model comparison with cross validation via \fct{sdmTMB\_cv}.
We will do 10-fold cross validation with the folds constructed randomly.
We will set the seed each time to ensure the folds are consistent.
Using the \code{fold\_ids} argument, we could supply our own folds and conduct spatially blocked cross validation \citep[][]{roberts2017}.
If we set a parallel plan with the \pkg{future} package \citep{bengtsson2021future}, our folds will be fit in parallel.

@@

where $n$ represents the number of left-out data points, $\ell$ represents the log likelihood, $y_i$ represents left-out data point $i$, and $\widehat{y_i}$ represents the prediction for left-out data point $i$ with the fixed effects at their MLE and random effects at their Empirical Bayes estimate.

@@

\sloppy To visualize our model, we can make predictions with the \fct{predict} method (\code{?predict.sdmTMB}) and optionally use the \code{newdata} argument to predict on a new data frame containing locations and all predictor columns.
Here, we will predict on a 2 \(\times\) 2 km grid (\code{qcs\_grid}) that covers the entire region of interest so we can visualize the predictions spatially.
The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict.  Some covariates might be fixed at a specified value for all predictions, such that we are predicting the expected value for samples conditional on those specified values.
We call these covariates \emph{detectability} or \emph{catchability} covariates, given the model is predicting the target variable while controlling for the additional influence of these detectability covariates that was estimated in fitted data \cite{thorson2019}.
The output of \fct{predict} is a data frame containing overall estimates in link space (\code{est}), estimates from the non-random-field components (\code{est\_non\_rf}; here, intercept and depth), and estimates from the individual random field components (here, \code{omega\_s}---the spatial field).
% The overall predictions combining fixed and random effects are plug-in predictions.
We can plot these with \fct{geom\_raster} from the \pkg{ggplot2} package (Figure~\ref{fig:pcod-predict-maps}).

@@

We can then specify our model. We include an offset for the effort variable log area swept such that we are effectively modelling density and our predictions will be for an area swept of 1 km\textsuperscript{2}.

Our model can be written as
\[
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
\exp \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
O_{\bm{s},t} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \gamma_t +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t}
\right),\\
\bm{\omega} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_\omega \right),\\
\bm{\epsilon}_{t} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_{\epsilon} \right),
\end{aligned}
\]
where \(\bm{\beta}\) are coefficients associated with the main effects, \(O_{\bm{s},t}\) represents the offset (here, log area swept), \(\gamma_t\) represents the time-varying coefficients, \(\omega_{\bm{s}}\) is the value from a spatial field (representing constant latent spatial effects) and \(\epsilon_{\bm{s},t}\) is a value from a spatiotemporal field  (representing latent spatial effects that vary by year).
The temporally varying intercepts \(\gamma_t\) are modeled as a stationary AR(1) process,
\[
\begin{aligned}
  \gamma_{t=1} &\sim \operatorname{Normal} \left(0, \sigma^2_{\gamma} \right),\\
  \gamma_{t>1} &\sim \operatorname{Normal} \left(\rho_\gamma\gamma_{t-1}, \sqrt{1 - \rho_\gamma^2} \sigma^2_{\gamma} \right),
\end{aligned}
\]
where \(\rho_\gamma\) represents the correlation between intercepts \(\gamma\) at time $t-1$ and $t$.
We include the argument \code{extra\_time}, which represents all time steps to include in the latent process, such that any autoregressive processes represent annual time steps as opposed to the irregular years of data collection.

@@

We can explore four alternative families that may better represent the data.
Each alternative family uses a delta model formulation as described in Section \ref{sec:delta}.
Note that we need to add an extra Newton optimization step to the delta-lognormal model to achieve suitable convergence (discovered via a call to \code{sanity(fit\_dl)}).
% Note that we choose to omit the spatiotemporal random fields from the binomial model component (\code{spatiotemporal = list("off", "iid")}) because initial experimentation indicated this random field was collapsing to zero indicating minimal changes in encounter spatial distribution from year to year.

@@

We next test two additional model formulations: making the spatial correlation isotropic (the default) instead of anisotropic, and structuring the positive model component spatiotemporal random fields as AR(1) to allow spatiotemporal patterns to partially persist from year to year.

The AR(1) fields can be represented as:
\[
\begin{aligned}
\bm{\epsilon}_{t=1} &\sim \operatorname{MVNormal} (\bm{0}, \bm{\Sigma}_{\epsilon}),\\
\bm{\epsilon}_{t>1} &= \rho \bm{\epsilon}_{t-1} + \sqrt{1 - \rho^2} \bm{\delta}_{t},  \:
\bm{\delta}_{t} \sim \operatorname{MVNormal} \left(\bm{0}, \bm{\Sigma}_{\epsilon} \right),
\end{aligned}
\]
where $\rho$ represents the estimated autoregressive parameter allowing the spatial field at time $t$ to be correlated with the spatial field at time $t-1$ with deviations created by $\bm{\delta}_t$, which are themselves independent random fields each year.
This is equivalent to a separable model GMRF with precision arising from the kronecker product of the spatial precision and an AR1 temporal precision.

@@

<<dog-residuals, fig.cap="Simulation-based randomized quantile residuals from the \\pkg{DHARMa} package. The statistical test is a two-sided Kolmogorov-Smirnov test of the null hypothesis that the residuals are drawn from a uniform(0, 1) distribution.", fig.asp=0.68, echo=TRUE>>=
set.seed(1)
s <- simulate(fit, nsim = 500, type = "mle-mvn")
dharma_residuals(s, fit)
@

>>>>

<<check-version, echo=FALSE, cache=FALSE>>=
if (packageVersion("sdmTMB") < "0.6.0") {
  stop("Please install a version of sdmTMB >= 0.6.0.", call. = FALSE)
}
@

@@

Here, we introduce the \R\ package \sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using \TMB\ for model fitting and \pkg{fmesher} or \INLA\ to set up SPDE matrices.
Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
Many individual features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date this full suite of useful features has not been integrated into a single package.
One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \TMB), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and approximated in discrete time-intervals.

@@

A GLMM with spatial Gaussian random fields can be written as
\[
\begin{aligned}
\mathbb{E}[y_{\bm{s}}] &= \mu_{\bm{s}},\\
\mu_{\bm{s}} &=
g^{-1} \left( \eta_{\bm{s}} \right),\\
\eta_{\bm{s}} &= \bm{X}_{\bm{s}} \bm{\beta} + \omega_{\bm{s}},\\
\end{aligned}
\]
where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in space $\bm{s}$ is defined as the mean $\mu_{\bm{s}}$.
That mean $\mu_{\bm{s}}$ is the result of an inverse link function $g^{-1}$ applied to a linear predictor $\eta_{\bm{s}}$.
In this case, that linear predictor is the combination of a model matrix $\bm{X}_{\bm{s}}$ multiplied by a vector of coefficients $\bm{\beta}$ and a value from a spatial random field $\omega_{\bm{s}}$.
This spatial random field represents the effect of latent spatial variables that are not otherwise accounted for in the model.
Alternatively, $\omega_{\bm{s}}$ can be thought of representing spatially correlated noise arising from unmodeled processes.
More simply, the vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) represents a ``wiggly'' surface with an expected value of zero that is added to the linear predictor in link space (e.g., Figure~\ref{fig:matern-range}).
The vector $\bm{\omega}$ is assumed drawn from a multivariate normal distribution with a covariance matrix $\boldsymbol{\Sigma}_\omega$,
\[
\boldsymbol{\omega} \sim \operatorname{MVNormal}
  \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),
\]
constrained by some function that defines the rate at which spatial covariance decays with distance.

@@

Various covariance functions are possible, but a popular and flexible one is the Mat\'ern \citep{whittle1954, matern1986} (Figure~\ref{fig:matern-range}).
If we define $\lVert h \rVert$ as the Euclidean distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the Mat\'ern covariance $\Phi$ as
\[
\Phi\left( \bm{s}_j, \bm{s}_k \right) =
  \frac{\sigma_m^2}{2^{\nu - 1}\Gamma(\nu)}
  \left(\kappa \lVert h \rVert \right)^\nu K_\nu
  \left( \kappa \lVert h \rVert \right).
\]
The parameter $\sigma_m^2$ is the marginal variance (magnitude of the random field ``wiggles''), $\Gamma$ represents the Gamma function, $K_\nu$ represents the modified Bessel function of the second kind, and $\kappa$ represents the spatial decorrelation rate.
The parameter $\nu$ controls the smoothness of the covariance function.
In practice, $\nu$ is challenging to estimate and herein is fixed at $\nu = 1$ \citep{lindgren2011}.
A more interpretable derived parameter than the spatial decorrelation rate $\kappa$ is the spatial range---a distance at which two points are effectively independent.
A common definition is $\sqrt{8 \nu} / \kappa$ (so if $\nu = 1$, $\mathrm{range} = \sqrt{8} / \kappa$), which has the empirically derived property of the distance at which correlation decays to $\approx 0.13$ \citep{lindgren2011} (Figure~\ref{fig:matern-range}c).

@@

In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or adding random intercepts by grouping factors (Figure~\ref{fig:diagram}).
All these components are additive in link space.
Adopting the notation ``main'' for main effects, ``tvc'' for time-varying coefficients, and ``svc'' for spatially varying coefficients, a more complex model could be written as
\[
\begin{aligned}
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma}_t +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\alpha_g +
O_{\bm{s},t} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right).
\end{aligned}
\]
where the $\bm{X}$ represent model matrices, $\bm{\gamma}_t$ represents a vector of coefficients that are constrained to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a random field, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O_{\bm{s},t}$ represents an offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}).
We demonstrate these model components in Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}.
% TODO: add MVN(0, sigma)s

@@

So far, we have described models with one linear predictor and one family (common terminology in \R\ packages for the combination of an observation likelihood and link).
Frequently, data are better represented with two-part ``delta'' or ``hurdle'' models, which include linear predictors and observation distributions for two processes: zero vs.\ non-zero values and positive values, respectively \citep{aitchison1955}.
Here, we describe two types of delta models, dropping the space and time subscripts for simplicity.
% and use $\bm{X_1} \bm{\beta_1}$ and $\bm{X_2} \bm{\beta_2}$ to refer to generic model matrices and parameter vectors for the first and second model components.

@@

Once the user makes a call to \fct{sdmTMB}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
If needed, data structures required to implement penalized smoothers are constructed with \fct{smooth2random} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
\fct{sdmTMB} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} \citep{Matrix} as an interface to the \pkg{Eigen} library in \proglang{C++}.
% TODO Jim:  I recommend adding a citation to package Matrix ( Bates D, Maechler M, Jagan M (2023). _Matrix: Sparse and Dense Matrix Classes and Methods_. R package version 1.6-4, <https://CRAN.R-project.org/package=Matrix>.)), as central dependency of TMB, given the detailed description of the dependency stack. If really motivated, you could also add the reference for Eigen (Guennebaud G, Jacob B, et al. (2010). “Eigen v3.” http://eigen.tuxfamily.org), although it's less clear what citation to use.
The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \Rev{A6} \fct{stats::optimHess} \citep{gay1990, r2021}.
Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
it is also possible to apply an ``epsilon'' estimator \citep{thorson2016bias}, which corrects for bias arising from the variance and skewness caused by non-linear transformations of random effects (see Section \ref{index-bias-correct} for an example).
Standard errors on all parameters and derived quantities---including those involving random effects---are calculated using the generalized delta method \citep{kristensen2016, zheng2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to sample from the joint posterior distribution for Bayesian inference.

@@

Prediction on fitted or new data is accomplished with \fct{predict} (\code{?predict.sdmTMB}).
In this paragraph, we include relevant \fct{predict} arguments in parentheses.
The predict method is flexible and includes the option to specify a new data frame (\code{newdata}),
whether to return predictions on the link or response scale (\code{type}),
whether to return standard errors (\code{se\_fit},
which can be slow if conditioned on random fields),
whether to condition on the random fields (\code{re\_form}),
whether to condition on the random intercepts (\code{re\_form\_iid}),
which delta model linear predictor to use (\code{model}),
whether to take draws from the joint parameter precision matrix (\code{nsim}),
and whether to use MCMC samples from a \pkg{tmbstan} model fit (\code{mcmc\_samples}).

A variety of model evaluation tools are available.
A \fct{residuals} method calculates various types of residuals.
The default is a form of randomized quantile \citep{dunn1996} or probability integral transform \citep{smith1985} residuals.
New observations can be simulated from a fitted model with \fct{simulate} or new data can be simulated without a fitted model (``de novo'') with \fct{sdmTMB\_simulate}.
\fct{sdmTMB\_cv} facilitates cross validation.

@@

<<remotes, eval=FALSE, echo=TRUE>>=
install.packages("pak")
pak::pkg_install("pbs-assess/sdmTMB", dependencies = TRUE)
@

@@

Users willing to replace the default \R\ \pkg{BLAS} \citep{blas}
% and \pkg{LAPACK} \citep{lapack}
library with an optimized version \citep[e.g., \pkg{openBLAS};][]{openblas}
can expect up to an order of magnitude increase in model fitting speed.
Suggestions are included in the package readme file

@@

We can visualize the mesh object with the associated plotting method (Figure~\ref{fig:dog-binomial-mesh2}).
Our mesh has 563 (\code{mesh\_pcod2\$mesh\$n}) vertices.
Mesh complexity has a large influence on the speed of fitting these models.

@@

<<morans, echo=TRUE>>=
inv_dist_matrix <- 1 / as.matrix(dist(pcod[,c("X", "Y"), ]))
diag(inv_dist_matrix) <- 0
set.seed(1)
r_bin <- residuals(fit_bin, type = "mle-mvn")
set.seed(1)
r_bin_rf <- residuals(fit_bin_rf, type = "mle-mvn")
ape::Moran.I(r_bin, weight = inv_dist_matrix)$p.value
ape::Moran.I(r_bin_rf, weight = inv_dist_matrix)$p.value
@

% We omit example code for Moran's I calculation for brevity but include it in the code supplement.
We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but a lack of evidence for spatial correlation after including a random field (p $>$ 0.01) suggesting that residual spatial autocorrelation is alleviated by including the random field.
We note that the specific p-value is dependent on the seed due to the randomization in the randomized quantile residuals \citep{dunn1996}.

@@

\Rev{A1}We note that caution is warranted in performing model selection via marginal AIC for models involving penalized smoothing (spatial or otherwise) \citep[e.g.][]{greven2010, safken2021}.
In marginal AIC calculation, the degrees of freedom is based on the number of fixed effects and does not account for the degree of random effect penalization as conditional AIC would.
Methods to estimate effective degrees of freedom for similar models were recently demonstrated in \citet{thorson2024cAIC}, but are not yet included in \sdmTMB.

@@

As an alternative to AIC, we can conduct model comparison with cross validation.
\sdmTMB\ includes the helper function \fct{sdmTMB\_cv} to facilitate this.
We will do 10-fold cross validation with the folds constructed randomly.
We will set the seed each time to ensure the folds are consistent.
Using the \code{fold\_ids} argument, we could supply our own folds and conduct spatially blocked cross validation \citep[][]{roberts2017}.
If we set a parallel plan with the \pkg{future} package \citep{bengtsson2021future}, our folds will be fit in parallel.

@@

where $n$ represents the number of left-out data points, $\ell$ represents the log likelihood, $y_i$ represents left-out data point $i$, and $\widehat{y_i}$ represents the prediction for left-out data point $i$ with the fixed effects at their MLEs and random effects at their empirical Bayes estimates.

@@

\sloppy To visualize our model, we can make predictions with the \fct{predict} method (\code{?predict.sdmTMB}) and optionally use the \code{newdata} argument to predict on a new data frame containing locations and all predictor columns.
Here, we will predict on a 2 \(\times\) 2 km grid (\code{qcs\_grid}) that covers the entire region of interest so we can visualize the predictions spatially.
The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict.
Some covariates might be fixed at a specified value for all predictions, such that we are predicting the expected value for samples conditional on those specified values.
In the context of fish or animal surveys, these are sometimes called \emph{detectability} or \emph{catchability} covariates, given the model is predicting the target variable while controlling for the additional influence of these detectability covariates \cite{thorson2019}.
The output of \fct{predict} is a data frame containing overall estimates in link space (\code{est}), estimates from the non-random-field components (\code{est\_non\_rf}; here, intercept and depth), and estimates from the individual random field components (here, \code{omega\_s}---the spatial field).
% The overall predictions combining fixed and random effects are plug-in predictions.
We can plot these with \fct{geom\_raster} or \fct{geom\_tile}  from the \pkg{ggplot2} \citep{ggplot2} package (Figure~\ref{fig:pcod-predict-maps}).

@@

We can then specify our model. We include an offset \citep[][p.~206]{mccullagh1989} for the effort variable log area swept such that we are effectively modelling density and our predictions will be for an area swept of 1 km\textsuperscript{2}.

Our model can be written as
\[
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
\exp \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
O_{\bm{s},t} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \gamma_t +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t}
\right),\\
\bm{\omega} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_\omega \right),\\
\bm{\epsilon}_{t} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_{\epsilon} \right),
\end{aligned}
\]
where \(\bm{\beta}\) are coefficients associated with the main effects, \(O_{\bm{s},t}\) represents the offset (here, log area swept), \(\gamma_t\) represents the time-varying coefficients, \(\omega_{\bm{s}}\) is the value from a spatial field (representing constant latent spatial effects) and \(\epsilon_{\bm{s},t}\) is a value from a spatiotemporal field  (representing latent spatial effects that vary by year).
The temporally varying intercepts \(\gamma_t\) are modeled as a stationary AR(1) process,
\[
\begin{aligned}
  \gamma_{t=1} &\sim \operatorname{Normal} \left(0, \sigma^2_{\gamma} \right),\\
  \gamma_{t>1} &\sim \operatorname{Normal} \left(\rho_\gamma\gamma_{t-1}, \sqrt{1 - \rho_\gamma^2} \sigma^2_{\gamma} \right),
\end{aligned}
\]
where \(\rho_\gamma\) represents the correlation between intercepts \(\gamma\) at time $t-1$ and $t$ and $\sigma^2_{\gamma}$ represents the marginal variance of this process.
We include the argument \code{extra\_time}, which represents all time steps to include in the latent process (and all time steps for which we may wish to predict), such that autoregressive processes are applied to equally spaced annual time steps.

@@

We next explore four alternative families that may better represent the data.
Each alternative family uses a delta model formulation as described in Section \ref{sec:delta}.

@@

We next test two additional model formulations: making the spatial correlation isotropic (the default) instead of anisotropic, and structuring the spatiotemporal random fields as AR(1) to allow spatiotemporal patterns to partially persist from year to year.

The AR(1) fields can be represented as
\[
\begin{aligned}
\bm{\epsilon}_{t=1} &\sim \operatorname{MVNormal} (\bm{0}, \bm{\Sigma}_{\epsilon}),\\
\bm{\epsilon}_{t>1} &= \rho \bm{\epsilon}_{t-1} + \sqrt{1 - \rho^2} \bm{\delta}_{t},  \:
\bm{\delta}_{t} \sim \operatorname{MVNormal} \left(\bm{0}, \bm{\Sigma}_{\epsilon} \right),
\end{aligned}
\]
where $\rho$ represents the estimated autoregressive parameter allowing the spatial field at time $t$ to be correlated with the spatial field at time $t-1$ with deviations created by $\bm{\delta}_t$, which are themselves independent random fields each year.
This is equivalent to a separable model GMRF with precision arising from the Kronecker product of the spatial precision and an AR(1) temporal precision.

@@

<<dog-residuals, fig.cap="Simulation-based randomized quantile residuals from the \\pkg{DHARMa} package. The statistical test is a two-sided Kolmogorov-Smirnov test of the null hypothesis that the residuals are drawn from a uniform(0, 1) distribution.", fig.asp=0.68, echo=TRUE, out.width="4in">>=
set.seed(42)
s <- simulate(fit, nsim = 500, type = "mle-mvn")
dharma_residuals(s, fit, test_uniformity = TRUE)
@

The quantile-quantile plot suggests that under the model assumptions, the (transformed) residuals are reasonably consistent with an independent uniform(0, 1) distribution.
